Appendix S3. The relative importance of evolution in populations and communities

### 2025-09-29 manuscript/Appendix_S2.Rmd
Rename Appendix code files

---
title: "Appendix S2. Evolutionary drivers of species and community dynamics"
author:
  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
library(ggplot2)
library(tidyr)
library(coda)
library(patchwork)
library(jtools)
library(bayesplot)
library(reshape2)
library(ecoevor)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

>>>>

---
title: "Appendix S2. Evolutionary drivers of species and community dynamics"
author:
- "Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route
  de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]"
- "Ruben J. Hermann^[University of Konstanz, Universitätsstraße 10, 78464 Konstanz,
  Germany, ruben.hermann@posteo.de]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
library(ggplot2)
library(tidyr)
library(coda)
library(patchwork)
library(jtools)
library(bayesplot)
bayesplot_theme_set(new = theme_default())
library(reshape2)
library(ecoevor)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

### 2025-09-29 manuscript/Appendix_S3.Rmd
Rename Appendix code files

---
title: "Appendix S3. The relative importance of evolution in populations and communities"
author:
  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(ecoevor)
library(ggplot2)
library(reshape2)
library(Hmsc)
library(bayesplot)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

@@

@@

```{r warning=FALSE}
VP1 <- computeVariancePartitioning(m.1.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.1.sample,VP1,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2} Simulated (dark blue points) population size over time for $h^2 = 0$, with posterior predicted mean (light blue points) and 95% highest density intervals (light blue bars)."}
# prepare for plot
pred2 <- predict(m.1.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```

@@

m.2.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m2.post.hmsc <- convertToCodaObject(m.2.sample)
```
```{r}
m2.post.hmsc <- convertToCodaObject(m.2.sample)
summary(m2.post.hmsc$Beta)
bayesplot::mcmc_trace(m2.post.hmsc$Beta)
bayesplot::mcmc_areas(m2.post.hmsc$Beta,area_method = c("equal height"))
```
```{r}
Gradient <- constructGradient(m.2.sample,focalVariable="dx1",non.focalVariables=list(E=list(2),Esq=list(2)),ngrid=59)
# Esq is manually constructed as gradient-produced E^2
Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
predY <- predict(m.2.sample,XData=Gradient$XDataNew,expected=TRUE)
plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="dx1_t",ylab="predicted N_t+1")
```

>>>>

---
title: "Appendix S3. The relative importance of evolution in populations and communities"
author:
- "Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route
  de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]"
- "Ruben J. Hermann^[University of Konstanz, Universitätsstraße 10, 78464 Konstanz,
  Germany, ruben.hermann@posteo.de]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(ecoevor)
library(ggplot2)
library(reshape2)
library(Hmsc)
library(bayesplot)
bayesplot_theme_set(new = theme_default())
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

@@

set.seed(10403)

@@

```{r, warning=FALSE,message=FALSE,results = 'hide'}
VP1 <- computeVariancePartitioning(m.1.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.1.sample,VP1,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2} Simulated (dark blue points) population size over time for $h^2 = 0$, with posterior predicted mean (light blue points) and 95% highest density intervals (light blue bars).", warning=FALSE}
# prepare for plot
pred2 <- predict(m.1.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```

@@

m.2.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m2.post.hmsc <- convertToCodaObject(m.2.sample)
```

```{r}
m2.post.hmsc <- convertToCodaObject(m.2.sample)
summary(m2.post.hmsc$Beta)
bayesplot::mcmc_trace(m2.post.hmsc$Beta)
bayesplot::mcmc_areas(m2.post.hmsc$Beta,area_method = c("equal height"))
```
```{r}
Gradient <- constructGradient(m.2.sample,focalVariable="dx1",non.focalVariables=list(E=list(2),Esq=list(2)),ngrid=59)
# Esq is manually constructed as gradient-produced E^2
Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
predY <- predict(m.2.sample,XData=Gradient$XDataNew,expected=TRUE)
plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="dx1_t",ylab="predicted N_t+1")
```
